---
title: "Introduction to R and RStudio"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "2021/01/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: inverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

## Pre-workshop tasks:

.big-text-font[
1. Download [R](https://www.r-project.org/) version 4.0.3
2. Download [RStudio](https://rstudio.com/products/rstudio/download/) version >= 1.3.1
3. View and download workshop materials at [Google Drive](https://drive.google.com/drive/folders/1f_zDdTaQ2uykCvLsDZIxz0y6f28DK1pn?usp=sharing) 
]

## Post-workshop

.big-text-font[
1. Post-workshop survey will be sent out
2. Recordings will be made available shortly after
]

---

class: center

# Outline

| Time              | Topic                          |
|-------------------|--------------------------------|
| 11:00am - 11:15am | Overview of R & RStudio        |
| 11:15am - 11:30am | Objects and functions          |
| 11:30am - 12:00pm | Data types and data structures |
| 12:00pm - 12:05pm | Break                          |
| 12:05pm - 12:30pm | Data Management                |
| 12:30pm - 12:55pm | Dataset demo (with plotting)   |
| 12:55pm - 1:00pm  | Questions                      |

---

# Learning Objectives

.big-text-font[
1. **Programming**: learn R programming syntax to interact with your data in complex and useful ways

2. **Data Management**: learn how to represent and manipulate data using R data structures

3. **Exploratory Data Analysis**: learn how to run descriptive statistics and generate visualizations of your data

4. **Reproducible research**: learn how to use R and RStudio to manage your projects and analyses in a reproducible manner
]

---

# Slide structure

.big-text-font[
- Slides will be a mix of text, images, and code
- For code examples, look for a light-yellow box
- R comments are prefixed by `#`
    - Comments are useful notes about the R code
    - Comments are not R code
- R code is not prefixed by anything
- R output is prefixed by `#>`
    - Output is displayed in the console pane
]

.code-bg[
```{r sample-R-code, collapse=TRUE}
# This is a comment
sum(1, 2, 3, 4, 5)
```
]

---

# Workshop structure

.big-text-font[
1. Lecture
2. Short demos in RStudio (when applicable)
3. Checkpoint Questions
  - Quick review questions after a topic/section 
  - Volunteers to answer and answer question
4. Live coding (last ~ 30 minutes)
]
---

class: inverse, center, middle

# Overview of R

---

## What is R?

.big-text-font[
R is a programming language with strengths in data management, data visualization, and statistics. 	
]

---

## Why R?

.big-text-font[
-	Free
- Welcoming community (IMO)
    - Lots of R bloggers, twitter presence, community groups (RLadies, local groups)
- Open source (non-proprietary)
- Active community development to solve problems
    - If you are trying to solve a problem, someone else has probably written a R solution that you could use
- Flexible and extensible
    - Dashboards 
    - Automate periodic reports
    - Develop novel statistical tools
]

---

## Why not R?

.big-text-font[
- Other software already suits your needs.
- Legacy systems:
    - Highly common to have years or decades of code and programs written in other languages. 
    - Cheaper to maintain existing systems than to invest the resources and human hours to convert to R.
- High investment of time and energy to learn and be productive in R
- You collaborate with others who do not use R
]

---

class: inverse, middle, center

# Overview of RStudio

---

## What is RStudio?

.big-text-font[
RStudio is a software tool for R that provides utilities for writing R code, data analysis, and project management
]

--

### Some features
.big-text-font[
- Code completion
- Syntax highlighting
- Integrated help and documentation 
- Develop and test interactive graphics
- Author reports, books, slides
]

---

# Navigating RStudio

.big-text-font[
- Rstudio can be divided into different panes. Below are the most important panes.

**Console**
  - Area to write R code
  - Meant for interactive use and quick checks

**Script**
  - Another area to write R code
  - Similar to a Word Document but for code
  - Analogous to Stata do files, SPSS and SAS syntax files
]

---

# Navigating RStudio

.big-text-font[
**Environment**
  - Area to view and inspect all your objects and data

**Files**
  - Area to navigate your files
  - Can open up R and other supported files directly
]
---

# Navigating RStudio

.big-text-font[
**Packages**
  - Lists all downloaded packages
  - Able to download and update packages
  - Packages are user-written R code to solve a specific problem
  - Packages are analogous to Stata user-written functions, SAS and SPSS macros
  
**Help**
  - Contains documentation for anything R-related
  - Get help by typing in script or console: `help("topic")` or `?topic`
]
