---
title: "Introduction to R and RStudio"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "2021/01/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```



## Pre-workshop tasks:

1. Download [R](https://www.r-project.org/) version 4.0.3
2. Download [RStudio](https://rstudio.com/products/rstudio/download/) version >= 1.3.1
3. View and download [workshop repository](https://github.com/antchau/workshops/tree/master/intro-base-r) 


## Post-workshop
1. Post-workshop survey will be sent out
2. Recordings will be made available shortly after

---

class: center

## Outline

| Time              | Topic                          |
|-------------------|--------------------------------|
| 11:00am - 11:15am | Overview of R & RStudio        |
| 11:15am - 11:30am | Objects and functions          |
| 11:30am - 12:00pm | Data types and data structures |
| 12:00pm - 12:05pm | Break                          |
| 12:05pm - 12:20pm | Data Management                |
| 12:20pm - 12:30pm | Plotting                       |
| 12:30pm - 12:50pm | Dataset walkthrough            |
| 12:50pm - 12:55pm | Beyond Base R                  |
| 12:55pm - 1:00pm  | Questions                      |

---

## Slide structure

.lots-of-text-font[
- Slides will be a mix of text, images, and code
- For code examples, look for a light-yellow box
- R comments are prefixed by `#`
    - Comments are useful notes about the R code
    - Comments are not R code
- R code is not prefixed by anything
- R output is prefixed by `#>`
    - Output is displayed in the console pane
]

<br>  

.code-bg[
```{r sample-R-code, collapse=TRUE}
# This is a comment
sum(1, 2, 3, 4, 5)
```
]

## Checkpoint Questions
- Quick review questions after a topic/section 
- Volunteers to answer and answer question

---

class: inverse, center, middle

# Overview of R

---

class: inverse

## What is R?

R is a programming language with strengths in data management, data visualization, and statistics. 	

---

class: inverse

## Why R?

- Open source
-	Free
- Huge number of external libraries (>= 16,890 on CRAN)
- Strong reproducible research tools
    - Help with issues such as:
        - Collaborating with others on the same project
        - Distributing your work to others
        - Scripting your whole workflow. Anything you do to data can and should be expressed in code 
- Flexible and extensible
    - Create novel visualizations for a scientific use
    - Automate periodic reports for your business, lab, team.
    - Example: Working on custom finance reporting: can write script to login, select, and download data, clean data, wrangle data into a composed report. 
- Develop statistical tools (Professor Daniel Gillen’s `RCT` package for clinical trials)
-  Welcoming community (IMO)
    - Lots of R bloggers, twitter presence, community groups (RLadies, local groups)
    
---

class: inverse

## Why not R?

- Other software already suits your needs.
- Legacy systems:
    - Highly common to have years or decades of code and programs written in other languages. 
    - It’s cheaper to maintain existing systems than to invest the resources and human hours to convert to R.
- Investment of time and energy to learn R. Takes a fair amount of upfront investment to be productive in R. 
- You collaborate with others who do not use R
- Memory issues - you require more computing resources to work with your data

---

class: inverse, middle, center

# Overview of RStudio

---

class: inverse

## What is RStudio?

RStudio is a software tool for R that provides utilities for software development, data analysis, and project management

--
### Some features
- Code completion
- Syntax highlighting
- Integrated help and documentation services
- Develop and test interactive graphics
- Author reports, books, slides

--

### Brief tour of RStudio

- Different panes: console, script, environment, files, plots, packages, help
- Getting help: `help("topic")` or `?topic`
- Packages
- Keyboard shortcuts





