---
title: "Intermediate dplyr"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "7/24/2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
library(dplyr)
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

# Beyond core dplyr

.big-text-font[
- `dplyr` provides many useful functions beyond the core verbs (select, filter, mutate, etc)
- Functions:
    - `if_else()`
    - `case_when()`
    - `across()`, `if_any`, `if_all`
    - `*_join()`functions
]

---

# if_else()

.big-text-font[
- `if_else()` allows you to create a new variable based on a condition
- Use `if_else()` within `mutate`
- General form: `ifelse(condition, value when condition is True, value when condition is False)`
]

---

# if_else() example

.code-bg[
```{r ifelse1, collapse=TRUE}
starwars <- dplyr::starwars
starwars %>% 
  # Each value in the films variable is a list. 
  # Compute the length of the films variable fir each character
  # If length is's greater than 1, character is recurring. 
  # Otherwise, they only have one appearance
  mutate(recurring = ifelse(length(films) > 1, "Yes", "No")) %>% 
  select(name, recurring, films) %>% 
  head(5)
```
]

---

# case_when()

.big-text-font[
- `case_when()` allows you to create a new variable based on many conditions
- Think of `case_when()` as a generalized version of `if_else`
- Use `case_when()` within `mutate`
- The conditions of `case_when()` are sequential - ie: if your data is matched by condition 1 and 2, the value of the new variable will the value specified for condition 1 since condition 1 was first
]

```
case_when(new_variable = 
            condition 1 ~ value when condition 1 is True, 
            condition 2 ~ value when condition 2 is True,
            more conditions ....,
            # if data doesn't match any of the above conditions, 
            assign the value to NA (missing)
            TRUE ~ NA)
```
---

# case_when() example

.code-bg[
```{r casewhen1, collapse=TRUE}
starwars %>% 
  mutate(tallness = case_when(
    height > 180 ~ "Tall",
    height >= 150 & height < 180 ~ "Average",
    height < 150 ~ "Short",
    TRUE ~ NA_character_)
    ) %>% 
  select(name, height, tallness) %>% 
  head(5)
```
]

---

# across()

.big-text-font[
- `across()` allows you to apply a function to multiple variables
- `across()` uses select helper functions to help you select variables efficiently
- `across()` reduces redundancy for similar operations performed to multiple variables
]

---

# across() example

.big-text-font[
- How to add a curve to all three test scores in one line of code? Use `across`
]

.code-bg[
```{r across1, collapse=TRUE}
# test scores
tests <- data.frame(test1 = round(runif(n=50, min = 0, max = 100)),
                    test2 = round(runif(n=50, min = 0, max = 100)),
                    test3 = round(runif(n=50, min = 0, max = 100))
)

tests %>% head(10)
                    
```
]

---

# across() example

.code-bg[
```{r across2, collapse=TRUE}
tests %>%
  # everything() selects all columns
  # .x refers to each individual column
  # for each column, add 10 (points)
  mutate(across(.cols = everything(), 
                .fns = ~ .x + 10)) %>% 
  head(10)
                    
```
]


---

# across() example

.code-bg[
```{r across3, collapse=TRUE}
starwars %>%
  # use logical expressions inside where() 
  # .x refers to each individual character column
  # for each column, capitalize the text
  mutate(across(.cols = where(is.character), 
                .fns = ~ toupper(.x))) %>% 
  select(where(is.character)) %>% 
  head(10)
                    
```
]

---

# if_any(), if_all()

.big-text-font[
- `if_any()` allows you to filter data if at least of the selected columns selected meet a condition
- `if_all()` allows you to filter data if all the selected columns meet a condition
]

---

# if_any() example

.code-bg[
```{r ifany1, collapse=TRUE}
tests %>%
  # only show rows with students who scored above 50 
  # on ANY of the three tests
  filter(if_any(.cols = starts_with("test"), .funs = ~ .x > 50)) %>% 
  head(10)
                    
```
]

---

# if_all() example

.code-bg[
```{r ifall1, collapse=TRUE}
tests %>%
  # only show rows with students who scored above 50 
  # on ALL three tests
  filter(if_all(.cols = starts_with("test"), .fns = ~ . > 50))
```
]

---

# join() functions

.big-text-font[
- `*_join()`functions allows you to combine multiple data frames (tables) together  
- There are many different types of joins: left, right, inner, full, semi, anti
- We can link tables together with **key variables**. **Key variables** uniquely identify an observation.
]

---

# Join examples

.code-bg[
```{r students, collapse=TRUE}
students <- data.frame(
  id = c(1, 3, 4, 5, 8),
  name = c("Anthony", "Beatrice", "Claudia", "Dan", "Eliza")
)
classes <- data.frame(
  student_id = c(1, 1, 2, 3, 4, 4),
  class = c("English", "Math", "Science", "History", "Math", "History")
)
students
classes
```
]

---

# Left Join

.big-text-font[
- Keep all rows in the left table regardless of a match
]

.code-bg[
```{r leftJoin, collapse=TRUE}
students %>% 
  # supply key variables in the by argument
  # notice how the names of the key variables do not have to be the same
  left_join(classes, by = c("id" = "student_id"), keep=TRUE)

# keep all rows in the "left" table (students)

# the class variable from the "right" table (classes) is appended
# rows are repeated for every match from the "right" table
# when there is no match in id, a missing value is returned
```
]

---

# Right Join

.big-text-font[
- Keep all rows in the right table regardless of matches
]

.code-bg[
```{r rightJoin, collapse=TRUE}
students %>% 
  right_join(classes, by = c("id" = "student_id"), keep=TRUE)

# keep all rows in the "right" table (classes) are kept

# left_join(students, classes, by = c("id" = "student_id")) is equivalent to
# right_join(classes, students, by = c("student_id" = "id"))

```
]

---

# Inner Join

.big-text-font[
- Keep only rows with matches
]

.code-bg[
```{r innerJoin, collapse=TRUE}
students %>% 
  inner_join(classes, by = c("id" = "student_id"), keep=TRUE)

# only keep rows with matches in both tables
# in other words, only ids 1, 3, and 4 are found in both tables
```
]

---

# Full Join

.big-text-font[
- Keep all rows from both tables regardless of matches
]

.code-bg[
```{r fullJoin, collapse=TRUE}
students %>% 
  full_join(classes, by = c("id" = "student_id"), keep=TRUE)

# return all rows from both tables
# fill values with NA where there are no matches
```
]

---

# Semi Join

.big-text-font[
- Keep all rows in left table that have a match in right table
- Use the semi join to see which rows will match
- If a lot of rows are returned from the semi join, then be happy!
]

.code-bg[
```{r semiJoin, collapse=TRUE}
students %>% 
  semi_join(classes, by = c("id" = "student_id"), keep=TRUE)

# notice that the all ids are in the classes table
```
]

---

# Anti Join

.big-text-font[
- Drop all rows in left table that have a match in right table
- Use the anti join to see which rows will not match
- If a lot of rows are returned from the anti join, investigate the cause for low matching.
]

.code-bg[
```{r antiJoin, collapse=TRUE}
students %>% 
  anti_join(classes, by = c("id" = "student_id"), keep=TRUE)

# notice that the ids are NOT in the classes table
```
]

---

# Issues with joining

.big-text-font[
- When joining works, you are rewarded with rich datasets.
- Be wary of data issues that could affect your join
    - Missing values in key variables
    - Data entry issues - errors, misspellings, inconsistencies
]

---

# Missing values in key variables

.big-text-font[
- `id` variable in `df2` has a missing value. 
- If we can't identify the observation, we can't join it!
]

.code-bg[
```{r missing-values-join-issue, collapse=TRUE}
df1 <- data.frame(id = c(1, 2, 3, 4), 
                  date = c("Jan 2020", "Jan 2020", "Mar 2020", "Mar 2020"))

df2 <- data.frame(id = c(1, 2, NA, 5), 
                  sex = c("male", "female", "male", "female"))
```
]

---

# Data entry issues

.big-text-font[
- Key variables need to be identical for accurate row matching
- Data manipulation is required to make the `name` variable in `covid_test` consistent with the `name` variable in `patients`
]

.code-bg[
```{r data-entry-join-issue, collapse=TRUE}
patients <- 
  data.frame(id = c(1:5),
             name = c("Anthony", "Rick", 
                      "Claudia", "Dan", "John"))

covid_tests <- data.frame(name = c("Anthony", 
                                   "RICKY",
                                   "Johnny",
                                   "Claudia C",
                                   "Daniel"),
                          covid_tested = c("Yes", "No", "Yes",
                                           "Yes", "No"))
```
]

---

# Summary

.big-text-font[
- Create new variables based on conditions with `if_else()` and `case_when()`
- Apply functions to multiple variables with `across()`
- Filter data if all or any selected variables meet the conditions with `if_any()` and `if_all()`
- Combine data from different tables with `*_join()` functions
]



