---
title: "Data Management with dplyr"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "7/24/2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
library(dplyr)
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

# What is data management?

.big-text-font[
- Data management is the practice of manipulating data into more useful, correct, and accurate forms
- Data cleaning is another synonymous term for data management
- Real data is never "clean" - most likely, the data needs heavy data cleaning before we can do any analysis and visualization
]

---

# What is dplyr

.big-text-font[
- `dplyr` is a package for general data management
- Straightforward syntax and intuitive names
- Works well in conjunction with `ggplot2`
- Core functions: `select`, `filter`, `mutate`, `arrange`, `summarise`
- Additional functions: `***_join` functions
]

---

# select()

.big-text-font[
- `select` selects specific columns from our data
- `select` is another way of expressing `df[,c("col1", "col2")]
]

---

# Aside: %>% pipe operator

.big-text-font[
- `dplyr` code benefits from the use of the pipe operator `%>%`
- Think of the pipe as chaining together expressions
- The pipe allow us to write less redundant code and condense long data manipulation workflows
- Analogy: In-N-Out assembly line
]

---

# select() example

.big-text-font[
- `starwars` dataset containing character information from dplyr
]

.code-bg[
```{r select1, collapse=TRUE}
starwars <- dplyr::starwars
colnames(starwars)
starwars %>% 
  select(name, height, species) %>% 
  head(5)
```
]

---

# select() helpers

.big-text-font[
- There are many support functions and utilities to help you select columns
- ie: select variables that match a pattern, excluding specific variables, select all variables
- See this help page: [select() helpers](https://dplyr.tidyverse.org/reference/select.html)
]

.code-bg[
```{r select2, collapse=TRUE}
starwars %>% 
  select(name, ends_with("color")) %>% 
  head(5)
```
]

---

# filter()

.big-text-font[
- `filter` filters data given a condition
- Use `filter` to reduce the size of your dataset
- Create expressions with logical operators (<, <=, ==, !=, >=, >) and boolean operators (!, |, &)
    - Show anyone with weight greater than 200 pounds : 
    `weight > 200`
    - Show data from Florida and California : 
    `state == "Florida & state == "California`
    - Exclude all United airlines flights: 
    `carrier != "United`
]

---

# filter() example

.code-bg[
```{r filter1, collapse=TRUE}
starwars %>% 
  filter(sex == "male") %>% 
  select(name, sex, everything()) %>% 
  head(3)

# we can combine dplyr functions
starwars %>% 
  filter(sex == "female") %>% 
  select(name, species) %>% 
  head(3)
```
]

---

# filter() example

.code-bg[
```{r filter2, collapse=TRUE}
# hair_color == "blond" & eye_color == "blue"
starwars %>% 
  filter(hair_color == "blond", eye_color == "blue") %>% 
  select(name, hair_color, eye_color, everything())
```
]

---

# filter() example

.code-bg[
```{r filter3, collapse=TRUE}
starwars %>% 
  filter(species == "Human" | species == "Droid") %>% 
  select(name, species) %>% 
  head(5)
```
]

---

# mutate()

.big-text-font[
- `mutate` creates new variables in the data
- Use `mutate` to create variables not in the original data
]

---

# mutate() example

.code-bg[
```{r mutate1, collapse=TRUE}
starwars %>% 
  mutate(height_ft = height / 30.48) %>% 
  select(name, height, height_ft) %>% 
  head(5)
```
]

---

# mutate() example

.code-bg[
```{r mutate2, collapse=TRUE}
# save data with new variables
starwars2 <- 
  starwars %>% 
  mutate(height_m = height / 100,
         bmi = mass / (height_m)^2) 

starwars2 %>% 
  select(name, height, height_m, mass, bmi)
```
]

---

# arrange()

.big-text-font[
- `arrange` orders rows by the values of variables in the data
- Use `arrange` to present a sorted version of your data 
]

---

# arrange() example

.code-bg[
```{r arrange1, collapse=TRUE}
starwars %>% 
  # ascending order by height
  arrange(height) %>% 
  select(name, height) %>% 
  head(8)
```
]

---

# arrange() example

.code-bg[
```{r arrange2, collapse=TRUE}
starwars %>% 
  # descending order by height
  arrange(desc(height)) %>% 
  select(name, height) %>% 
  head(8)
```
]

---

# arrange() example

.code-bg[
```{r arrange3, collapse=TRUE}
mtcars %>% 
  # descending order by number of cylinders and then ascending order by mass
  arrange(desc(cyl), mpg) %>% 
  select(cyl, mpg) %>% 
  head(8)
```
]

---

# summarise()

.big-text-font[
- `summarise` creates a new data frame with computed "summary" functions
- Use `summarise` to compute descriptive statistics and summary information about the data
]

---

# summarise() example

.code-bg[
```{r summarise1, collapse=TRUE}
starwars %>% 
  # remove missing values from height before computing height
  # notice the naming of the summary variable on the left hand side
  summarise(mean_height = mean(height, na.rm=TRUE))


starwars %>% 
  # n() computes the number of observations
  summarise(max_mass = max(height, na.rm=TRUE),
            n = n())
```
]


---

# Aside: group_by()

.big-text-font[
- Often, we would like to compute summary information for different groups
- Groups are like categories in a categorical variable. 
- `group_by()` is useful when used with `mutate` or `summarise` to create group level variables or summaries
- `mutate` example: Add a new variable that contains the mean weight for males only and the mean weight for females only
- `summarise` example: Compute a new data frame of the mean weight for males only and the mean weight for females only
]

---

# Group by summarise()

.code-bg[
```{r summarise2, collapse=TRUE}
starwars %>% 
  group_by(sex) %>% 
  summarise(mean_height = mean(height, na.rm=TRUE))
```
]

---

# Group by mutate() 

.code-bg[
```{r mutate3, collapse=TRUE}
starwars %>% 
  group_by(sex) %>% 
  mutate(mean_height = mean(height, na.rm=TRUE)) %>% 
  select(name, sex, height, mean_height) %>% 
  head(6)
```
]

---

# rename()

.big-text-font[
- `rename()` renames variables
- `rename_with()` can rename multiple variables with a function
]

---

# rename() example

.code-bg[
```{r rename, collapse=TRUE}
starwars2 <- 
  starwars %>% 
  # new_name = old_name
  rename(height_cm = height) 
  

colnames(starwars)

# notice height is now height_cm
colnames(starwars2)
```
]

---

# rename() example

.code-bg[
```{r rename2, collapse=TRUE}
starwars2 <- 
  starwars %>% 
  # rename_with looks at every variable name by default
  # Use ~ to signal that we are supplying a function
  # gsub replaces all underscores with dot
  # .x indicates a single variable name
  rename_with(~gsub(pattern = "_",  
                    replacement = ".", 
                    x = .x)) 
  
colnames(starwars)
# notice the _ replaced by .
colnames(starwars2)
```
]

---

# Recap 

.big-text-font[
- `dplyr` facilitates quick and useful data manipulation
- Select specific variables with `select`
- Filter data with `filter`
- Create new variables with `mutate`
- Order data with `arrange`
- Summarize data with `summarise`
- Use `group_by` to perform grouped operations
]


