---
title: "Packages"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "2021/01/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

```{r tables-html, eval=params$html_slides,include=FALSE}
packages_table_path <- "slides/packages/tables/popular-packages.txt"
```

```{r tables-pdf, eval=params$pdf_slides,include=FALSE}
packages_table_path <- "tables/popular-packages.txt"
```


# Packages

.big-text-font[
- **package**: user-written R code and functions to solve a specific problem
- Packages are submitted to **CRAN** (official repository for R packages)  
- R has a huge number of external packages (>= 17,000 on CRAN)
- Browse packages at [r-pkg.org](https://www.r-pkg.org/)
]

---

# Popular Packages

.center[
`r knitr::asis_output(paste(readLines(packages_table_path), collapse = '\n'))`
]

---

# Installing packages

.big-text-font[
- Install a package with `install.packages("package_name")`
- Normally, package installation is done once
- You can update a package within the Packages pane in RStudio or with `update.packages()`
]

.code-bg[
```{r install-package, collapse=TRUE, eval=FALSE}
install.packages("dplyr")
```
]

---

# Loading packages

- We can use the functions in a package by loading a package into our environment
- Load a package with `library("package_name")`

.code-bg[
```{r load-package, collapse=TRUE, eval=FALSE}
library(dplyr)
```
]

---

# Using functions from a package

- After loading a package into your environment, you can use the functions in the package
- You can use functions from a package just as usual: `function_name(arg1 = val1, arg2 = val2, ...)`

.code-bg[
```{r use-package-function, collapse=TRUE, message=FALSE}
library(dplyr)

df <- data.frame(a = c(1,2,3),
                 b = c(4,5,6),
                 c = c(7,8,9))

# select columns from a data frame
select(df, a, c)
```
]

---

# Using functions from a package

- Alternatively, use the `package_name::package_function()` syntax to be explicit which function the package belongs to

.code-bg[
```{r use-package-function-colon, collapse=TRUE, message=FALSE}
library(dplyr)
df <- data.frame(a = c(1,2,3),
                 b = c(4,5,6),
                 c = c(7,8,9))

# select columns from a data frame
dplyr::select(df, a, c)
```
]


