---
title: "Data Visualization with ggplot2"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "7/24/2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

# What is ggplot2?

.big-text-font[
- `ggplot2` is a R package for creating statistical and data graphics
- `ggplot2`'s approach to graphics is based on [The Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
- Mature package
- Powerful and extensible
]

---

# Grammar of Graphics
.big-text-font[
- Big idea: a visualization is constructed from many independent components
- We put together different components to create our desired visualization
- Components of a plot:
    - Data
    - Aesthetic mappings
    - Geometric objects
    - Scales
    - Facet specification
    - Statistical Transformation
    - Coordinate System
]

---

# US Midwest Demographics

.big-text-font[
- Let's use ggplot2 on a [dataset](https://ggplot2.tidyverse.org/reference/midwest.html) containing demographics information for the US Midwest from the 2000 Census
]

.code-bg[
```{r mpg, collapse=TRUE}
midwest <- ggplot2::midwest
midwest[1, c(1,2,3,4,5)]
dim(midwest)
colnames(midwest)
```
]

---
# Scatterplot example

.big-text-font[
- Suppose we wanted to know the relationship between the percent of people below poverty line and the percent of people college educated
- What would you expect the graph to look like?
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot, eval=FALSE}
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege) + 
  geom_point()
```
]

.right-plot[
```{r midwest-scatterplot-out, ref.label="midwest-scatterplot", echo=FALSE}
```
]

---

# Recap

.big-text-font[
- Intialize a plot with the `ggplot` function
- Specify our data source
- *Aesthetic* properties: choose which variables to use for x and y position
- *Geometric* object (geom): Specify the type of plot 
- We used the *point geom* `geom_point()` which produces a scatterplot
]

---

# Extending the plot

.big-text-font[
- Suppose we wanted to see how the relationship between the percent of people below poverty line and the percent of people college educated depends on if someone lives in a metropolitian area
- We can display this visually by assigning another aesthetic element to your desired variable
- Let's use the color aesthetic and map it to the `inmetro` variable
- List of common aesthetics: color, shape, size, line type, line size, transparency
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot2, eval=FALSE}
midwest$inmetro <- 
  factor(midwest$inmetro,
         levels = c("0", "1"),
         labels = c("No", "Yes"))
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege,
      color = inmetro) + #<< 
  geom_point() 
```
]

.right-plot[
```{r midwest-scatterplot2-out, ref.label="midwest-scatterplot2", echo=FALSE}
```
]

---

# Improving the scatterplot

.big-text-font[
- Let's make some adjustment to the formatting of the plot
- List of changes:
    - Add descriptive plot title
    - Make axis labels more descriptives
    - Change legend title
    - Add spacing between axis and axis labels 
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot3, eval=FALSE}
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege,
      color = inmetro) +
  geom_point() +
  labs(title = "% College Educated by % Adults below poverty line", #<<
       subtitle = "2000 US Midwest Census", #<<
       color = "In Metro Area", #<<
       x = "% Adults below poverty line", #<<
       y = "% College Educated") + #<< 
    theme( #<<
      axis.title.x = #<<
        element_text( #<<
          margin = margin(t = 20, r = 0, #<<
                          b = 0, l = 0)), #<<
      axis.title.y = #<<
        element_text( #<<
          margin = margin(t = 0, r = 20, #<<
                          b = 0, l = 0)) #<<
    )
```
]

.right-plot[
```{r midwest-scatterplot3-out, ref.label="midwest-scatterplot3", echo=FALSE}
```
]

---

# Histogram example

.big-text-font[
- Suppose we want to know the population density distribution for all counties
- A histogram can be used to view the distribution
- The histogram geom (`geom_histogram`) in `ggplot2` displays a histogram
]

---

# Histogram example

.left-code[
```{r midwest-histogram, eval=FALSE, warning=FALSE}
ggplot(midwest) +
  aes(x = popdensity) + 
  geom_histogram()
```
]

.right-plot[
```{r midwest-histogram-out, ref.label="midwest-histogram", echo=FALSE}
```
]

---

# Improving the histogram
.big-text-font[
- The x-axis label can be changed to be more descriptive 
- We can tinker with the binwidth (how wide each individual bar is) to display our desired plot 
- Increase the number of x-axis *breaks* (positions on the axis that are marked)
- Adjust the x-axis *limits* to only include the majority of the data
- Make the bars hollow and have an outline color for the bars 
]

---

# Improving the histogram

.left-code[
```{r midwest-histogram2, eval=FALSE}
ggplot(midwest) +
  aes(x = popdensity) + 
  geom_histogram(binwidth = 500,
                 fill = "white",
                 color = "purple") +
  scale_x_continuous(breaks = seq(0, 15000, 1500),
                     limits = c(0, 15000)) +
  labs(title = "Population Density of US Midwest Counties",
       x = "Population Density (person / unit area)") +
  theme_minimal()
```
]

.right-plot[
```{r midwest-histogram2-out, echo=FALSE, warning=FALSE, ref.label="midwest-histogram2"}
```
]

---
# Barplot example

.big-text-font[
- Now, let's see how many counties are in each state 
- A barplot can be used to view the distribution of a qualitative variable
- The bar geom (`geom_bar`) or the col geom (`geom_col`) in `ggplot2` displays a barplot
- The difference between `geom_bar` and `geom_col` is that `geom_bar` makes the height of the bar proportional to the number of each case in each group and `geom_col` makes the heights of the bars represent values in the data
]

---

# Barplot example

.left-code[
```{r midwest-barplot, eval=FALSE}
ggplot(midwest) +
  aes(x = state) + 
  geom_bar()
```
]

.right-plot[
```{r midwest-barplot-out, ref.label="midwest-barplot", echo=FALSE}
```
]

---

# Improving the barplot

.big-text-font[
- Order the barplot by count
- Flip coordinate system
- Make the bars hollow and have an outline color for the bars
]

---

# Improving the barplot

.left-code[
```{r midwest-barplot2, eval=FALSE}
# count number of counties in each state
county_count <- as.data.frame(table(midwest$state))
# rename column
colnames(county_count)[colnames(county_count) == 'Var1'] <- 'state'

ggplot(county_count) +
  aes(x = reorder(state, Freq),
      y = Freq) + 
  geom_col() +
  coord_flip() +
  labs(title = "Number of counties in each state",
       x = "State",
       y = "Count")
```
]

.right-plot[
```{r midwest-barplot2-out, ref.label="midwest-barplot2", echo=FALSE}
```
]

---

# Boxplot example

.big-text-font[
- Now, let's see the distribution of the percent of people below the poverty line for each state
- A boxplot can be used to view the distribution of a quantitative variable by a categorical variable
- The boxplot geom (`geom_boxplot`) in `ggplot2` displays a boxplot
]


---

# Boxplot example

.left-code[
```{r midwest-boxplot, eval=FALSE}
ggplot(midwest) +
  aes(x = state, 
      y = percbelowpoverty) + 
  geom_boxplot() +
  labs(title = "% population below poverty line",
       subtitle = "2000 US Census - Midwest States",
       y = "% below poverty")
```
]

.right-plot[
```{r midwest-boxplot-out, ref.label="midwest-boxplot", echo=FALSE}
```
]

---

# Summary

.big-text-font[
- `ggplot2` builds a plot by combining multiple components
- Plots need a **data** source as well as variables to map to the x and y positions
- **Aesthetics** are visual properties of the plot - what you can see on the plot. Common aesthetics are the x position, y position, color, shape, and size.
- The selection of a **geom** determines the type of plot
- The **scale** can be modified to change how the data maps over to aesthetic properties. There is one scale for each aesthetic property.
- The **coord** can be modified to change the position of objects relative to the plane of the plot
]

---

# Summary

.big-text-font[
- The **theme** controls the non-data components of the plot. Some examples: titles, labels, fonts, background, gridlines. See `?theme`.
- **Facets** specify how to display subsets of your data
- **Statistics** are transformations of your data that can be drawn on top of the data 
]

