---
title: "Data Visualization with ggplot2"
author: "Anthony Chau"
institute: "UCI Center for Statistical Consulting"
date: "7/24/2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

# What is ggplot2?

.big-text-font[
- `ggplot2` is a R package for creating statistical and data graphics
- `ggplot2`'s approach to graphics is based on [The Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
- Mature package
- Powerful and extensible
]

---

# Grammar of Graphics
.big-text-font[
- Big idea: a visualization is constructed from many independent components
- We put together different components to create our desired visualization
- Components of a plot:
    - Data
    - Aesthetic mappings
    - Geometric objects
    - Scales
    - Facet specification
    - Statistical Transformation
    - Coordinate System
]

---

# US Midwest Demographics

.big-text-font[
- Let's use ggplot2 on a [dataset](https://ggplot2.tidyverse.org/reference/midwest.html) containing demographics information for the US Midwest from the 2000 Census
]

.code-bg[
```{r mpg, collapse=TRUE}
midwest <- ggplot2::midwest
midwest[1, c(1,2,3,4,5)]
dim(midwest)
colnames(midwest)
```
]


---

# Scatterplots
.left-column[
- A scatterplot visualizes the relationship between two quantitative variables
- The data points are commonly represented as points.
]

.right-column[
```{r scatterplot, echo=FALSE}
x <- seq(1, 50) + runif(n = 50)
y <- seq(1, 50) + runif(n = 50, max = 5)
plot(x,y)
```
]

---

# Scatterplot example

.big-text-font[
- Suppose we wanted to know the relationship between the percent of people below poverty line and the percent of people college educated
- We can use a scatterplot to visualize the relationship since both variables are quantitative
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot, eval=FALSE}
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege) + 
  geom_point()
```
]

.right-plot[
```{r midwest-scatterplot-out, ref.label="midwest-scatterplot", echo=FALSE}
```
]

---

# Recap

.big-text-font[
- Intialize a plot with the `ggplot` function
- Specify our data source
- *Aesthetic* properties: choose which variables to use for x and y position
- *Geometric* object (geom): Specify the type of plot 
- We used the *point geom* `geom_point()` which produces a scatterplot
]

---

# Extending the plot

.big-text-font[
- Suppose we wanted to see how the relationship between the percent of people below poverty line and the percent of people college educated depends on if someone lives in a metropolitian area
- We can display this visually by assigning another aesthetic element to your desired variable
- Let's use the color aesthetic and map it to the `inmetro` variable
- List of common aesthetics: color, shape, size, line type, line size, transparency
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot2, eval=FALSE}
midwest$inmetro <- 
  factor(midwest$inmetro,
         levels = c("0", "1"),
         labels = c("No", "Yes"))
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege,
      color = inmetro) + #<< 
  geom_point() 
```
]

.right-plot[
```{r midwest-scatterplot2-out, ref.label="midwest-scatterplot2", echo=FALSE}
```
]

---

# Improving the scatterplot

.big-text-font[
- Let's make some adjustment to the formatting of the plot
- List of changes:
    - Add descriptive plot title
    - Make axis labels more descriptives
    - Change legend title
    - Add spacing between axis and axis labels 
]

---

# Scatterplot example

.left-code[
```{r midwest-scatterplot3, eval=FALSE}
ggplot(midwest) +
  aes(x = percadultpoverty, 
      y = percollege,
      color = inmetro) +
  geom_point() +
  labs(title = "% College Educated by % Adults below poverty line", #<<
       subtitle = "2000 US Midwest Census", #<<
       color = "In Metro Area", #<<
       x = "% Adults below poverty line", #<<
       y = "% College Educated") + #<< 
    theme( #<<
      axis.title.x = #<<
        element_text( #<<
          margin = margin(t = 20, r = 0, #<<
                          b = 0, l = 0)), #<<
      axis.title.y = #<<
        element_text( #<<
          margin = margin(t = 0, r = 20, #<<
                          b = 0, l = 0)) #<<
    )
```
]

.right-plot[
```{r midwest-scatterplot3-out, ref.label="midwest-scatterplot3", echo=FALSE}
```
]

---

# Histograms

.big-text-font[
- A histogram can be used to view the distribution for a single quantitative variable
- The histogram geom (`geom_histogram`) in `ggplot2` displays a histogram
]

---

# Histograms

.big-text-font[
- A histogram divides your data into equal sized bins and draws rectangular bars to represent each bin.
- For example, suppose you want to visualize the percentage of high school graduates. The histogram can have ten bins with the width of each bin being 10%
- The height of a bar represents how many times a value in a bin occurs
- The binwidth in a histogram is important for interpretation
]

---

# Histograms

.left-column[
- Binwidth = 10
]

.right-column[
```{r histogram1, echo=FALSE, out.width="100%"}
x <- c(seq(0, 50) + runif(n=51),
       seq(50, 70, by = 2),
       seq(70, 100) + runif(n=31))
hist(x)
```
]

---

# Histograms

.left-column[
- Binwidth = 20
]

.right-column[
```{r histogram2, echo=FALSE, out.width="100%"}
x <- 
  c(
    seq(0, 10, by = 2) + runif(n = 6),
    seq(10, 20, by = 2) + runif(n = 6),
    seq(20, 30, by = 2) + runif(n = 6),
    seq(30, 40, by = 2) + runif(n = 6),
    seq(40, 50, by = 2) + runif(n = 6),
    seq(50, 60, by = 5) + runif(n = 3),
    seq(60, 70, by = 2) + runif(n = 6),
    seq(70, 80, by = 5) + runif(n = 3),
    seq(80, 90, by = 1) + runif(n = 11),
    seq(90, 100, by = 2) + runif(n = 6)
    )
hist(x)
```
]

---

# Histogram example

.big-text-font[
- Suppose we want to know the distribution of population density for all counties
- We can use a histogram to visualize this distribution
- `ggplot2` will choose a default binwidth for you. Usually, you should change the binwidth
]

---

# Histogram example

.left-code[
```{r midwest-histogram, eval=FALSE, warning=FALSE}
ggplot(midwest) +
  aes(x = popdensity) + 
  geom_histogram()
```
]

.right-plot[
```{r midwest-histogram-out, ref.label="midwest-histogram", echo=FALSE}
```
]

---

# Improving the histogram
.big-text-font[
- The x-axis label can be changed to be more descriptive 
- We can tinker with the binwidth (how wide each individual bar is) to display our desired plot 
- Increase the number of x-axis *breaks* (positions on the axis that are marked)
- Adjust the x-axis *limits* to only include the majority of the data
- Make the bars hollow and have an outline color for the bars 
]

---

# Improving the histogram

.left-code[
```{r midwest-histogram2, eval=FALSE}
ggplot(midwest) +
  aes(x = popdensity) + 
  geom_histogram(binwidth = 500,
                 fill = "white",
                 color = "purple") +
  scale_x_continuous( #<<
    breaks = seq(0, 15000, 1500), #<<
    limits = c(0, 15000)) + #<<
  labs(title = "Population Density of US Midwest Counties", #<<
       x = "Population Density (person / unit area)") + #<<
  theme_minimal()
```
]

.right-plot[
```{r midwest-histogram2-out, echo=FALSE, warning=FALSE, ref.label="midwest-histogram2"}
```
]

---
# Barplots

.big-text-font[
- A barplot can be used to view the distribution of a categorical variable
- A barplot draws rectangular bars where the height represents some numerical quantity.
- This "numerical quantity" can be how many times a category occurs or the value of another quantitative variable for each category
- Barplots = categorical variables
- Histograms = quantitative variables
]

---

# Barplots

.big-text-font[
- Height of bar = Count of each unique category
    - Number of students enrolled at different universities
    - Categorical variable: "enrolled_university"
    - Count how many times "UCI", "UCLA", "USC" occur in the `enrolled_university` variable
- Height of bar = Value of another quantitative variable 
    - Mean age of students at different universities
    - No counting - display the value of a quantitative variable (`mean_age`) for each category
]

---

# Barplots

.left-column[
- The barplot displays how many people of each sex
]

.right-column[
```{r barplot1, echo=FALSE}
sex <- c(rep("Male", 5),
         rep("Female", 9),
         rep("Not reported", 3))

sex_cat <- c("Male", "Female", "Not reported")

sex <- data.frame(matrix(table(sex)))
sex$cat <- sex_cat
names(sex)[names(sex) == "matrix.table.sex.."] <- "count"

barplot(sex$count, names=sex$cat)
```
]

---

# Barplots

.left-column[
- The barplot displays the number of stores for each coffee chain
]

.right-column[
```{r barplot2, echo=FALSE}
stores <- data.frame(name = c("Starbucks", "Dunkin'", "Tim Horton's"),
                     num_stores = c(32660, 12871, 4846))

barplot(stores$num_stores, names=stores$name)
```
]

---

# Barplot example

.big-text-font[
- Now, let's see how many counties are in each state 
- Since state is categorical, we can use a barplot to display how many counties are in each state
- The bar geom (`geom_bar`) and the col geom (`geom_col`) in `ggplot2` displays a barplot
- The difference between `geom_bar` and `geom_col` is that `geom_bar` makes the height of the bar proportional to the number of each case in each group and `geom_col` makes the heights of the bars represent values in the data
- Knowing the above point, we need to use `geom_bar`
]

---

# Barplot example

.left-code[
```{r midwest-barplot, eval=FALSE}
ggplot(midwest) +
  aes(x = state) + 
  geom_bar()
```
]

.right-plot[
```{r midwest-barplot-out, ref.label="midwest-barplot", echo=FALSE}
```
]

---

# Improving the barplot

.big-text-font[
- Order the barplot by count
- Flip coordinate system
- Make the bars hollow and have an outline color for the bars
]

---

# Improving the barplot

.left-code[
```{r midwest-barplot2, eval=FALSE}
# count number of counties in each state
county_count <- 
  as.data.frame(
    table(midwest$state)
    )

county_count

# rename column
colnames(county_count)[colnames(county_count) == 'Var1'] <- 'state'

ggplot(county_count) +
  aes(x = reorder(state, Freq),
      y = Freq) + 
  geom_col() +
  coord_flip() + #<<
  labs(title = "Number of counties in each state",
       x = "State",
       y = "Count")
```
]

.right-plot[
```{r midwest-barplot2-out, ref.label="midwest-barplot2", echo=FALSE}
```
]

---

# Boxplots

.big-text-font[
- A boxplot can be used to view the distribution of a quantitative variable by a categorical variable
- A box with "whiskers" are drawn to show the distribution for each category 
- Often, we can read off the first quartile (Q1), median, and third quartile (Q3) from the boxplot.
- The **Interquartile range** (IQR) is the third quartile minus the first quartile: Q3-Q1
- The IQR is a measure of dispersion. A large IQR indicates that is greater spread
]

---

# Boxplots

.left-column[
- Top "Whisker": Q3 + 1.5 * IQR
- Top line of box: First Quartile
- Bold line: Median
- Bottom line of box: Third Quartile
- Bottom "Whisker": Q1 - 1.5 * IQR
- Additional points: outliers
]

.right-column[
```{r boxplot1, echo=FALSE, out.width="100%"}
iris <- iris
boxplot(Petal.Length ~ Species, data = iris)
```
]

---

# Boxplot example

.big-text-font[
- Now, let's see the distribution of the percent of people below the poverty line for each state
- A boxplot can be used to see how the percentage of people below the poverty line varies for each state
- The boxplot geom (`geom_boxplot`) in `ggplot2` displays a boxplot
]

---

# Boxplot example

.left-code[
```{r midwest-boxplot, eval=FALSE}
ggplot(midwest) +
  aes(x = state, 
      y = percbelowpoverty) + 
  geom_boxplot() +
  labs(title = "% population below poverty line",
       subtitle = "2000 US Census - Midwest States",
       y = "% below poverty")
```
]

.right-plot[
```{r midwest-boxplot-out, ref.label="midwest-boxplot", echo=FALSE}
```
]

---

# Summary

.big-text-font[
- `ggplot2` builds a plot by combining multiple components
- Plots need a **data** source as well as variables to map to the x and y positions
- **Aesthetics** are visual properties of the plot - what you can see on the plot. Common aesthetics are the x position, y position, color, shape, and size.
- The selection of a **geom** determines the type of plot
- The **scale** can be modified to change how the data maps over to aesthetic properties. There is one scale for each aesthetic property.
- The **coord** can be modified to change the position of objects relative to the plane of the plot
]

---

# Summary

.big-text-font[
- The **theme** controls the non-data components of the plot. Some examples: titles, labels, fonts, background, gridlines. See `?theme`.
- **Facets** specify how to display subsets of your data
- **Statistics** are transformations of your data that can be drawn on top of the data 
]

