---
title: "Data Interpretation"
author: "Anthony Chau"
date: "7/24/2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "style.css", "metropolis-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  html_slides: TRUE
  pdf_slides: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), 
                      fig.retina=2, 
                      out.width="100%",
                      comment = "#>")
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
.code-bg .remark-code {
  background-color: rgba(255, 210, 0, 0.2);
}
```

# Data Intrepretation

.big-text-font[
- After visualizing our data, it's useful to provide an interpretation of the visualization
- We'll focus on how to describe the relationship between two quantitative variables 
- Three ways to describe relationship: linearity, strength, and direction
]

---

# Linear relationships

.big-text-font[
- The relationship between two quantitative variables can be called *linearly related* if the distribution of the variables follow a straight line
- Now, it is rare for real data to have a perfectly linear relationship
- It's more likely that the relationship between two variablesroughly follows a straight line
- The opposite of a linear relationship is a *non-linear* relationship. For example, two variables may be quadratrically related.
]

---

# Perfect linearity

.left-code[
```{r perfect-linearity, eval=FALSE}
x <- c(1:50)
y <- c(1:50)
# plot points
plot(x, y)
# add 45 degree line
abline(a = 0, 
       b = 1)
```
]

.right-plot[
```{r perfect-linearity-out, ref.label="perfect-linearity", echo=FALSE}
```
]

---

# General Linearity

.left-code[
```{r general-linearity, eval=FALSE}
x <- c(1:50)
y <- c(1:50) + 
  rnorm(n = 50, 
        mean = 0, 
        sd = 4)
plot(x, y)
```
]

.right-plot[
```{r general-linearity-out, ref.label="general-linearity", echo=FALSE}
```
]

---

# Non-Linear Relationship

.left-code[
```{r non-linear, eval=FALSE}
x <- c(-50:50)
y <- x^2
plot(x, y)
```
]

.right-plot[
```{r non-linear-out, ref.label="non-linear", echo=FALSE}
```
]

---

# Strength of relationship

.big-text-font[
- We can also describe the strength of the relationship between two variables
- The strength of the relationship can be thought of as how closely two variables follow each other.
- If two variables line up exactly on a 45-degree line --> *perfect relationship*
- If two variables tend to cluster closely together --> *strong relationship*
- If two variables tend to spread out --> *weak relationship*
- If two variables make a horizontal line --> *no relationship*
]

---

# Strong relationship

.left-code[
```{r strong-relationship, eval=FALSE}
x <- c(1:50)
y <- c(1:50) + 
  rnorm(n = 50, 
        mean = 0, 
        sd = 2)
plot(x, y)
abline(a = 0, 
       b = 1, 
       col = "red")
```
]

.right-plot[
```{r strong-relationship-out, ref.label="strong-relationship", echo=FALSE}
```
]

---

# Weak relationship

.left-code[
```{r weak-relationship, eval=FALSE}
x <- c(1:50)
y <- c(1:50) + 
  rnorm(n = 50, 
        mean = 0, 
        sd = 13)
plot(x, y)
abline(a = 0, 
       b = 1, 
       col = "red")
```
]

.right-plot[
```{r weak-relationship-out, ref.label="weak-relationship", echo=FALSE}
```
]

---

# No relationship

.left-code[
```{r no-relationship, eval=FALSE}
x <- c(1:50)
y <- rep(15, 50)
plot(x, y)
abline(a = 15, 
       b = 0, 
       col = "red")
```
]

.right-plot[
```{r no-relationship-out, ref.label="no-relationship", echo=FALSE}
```
]

---

# Direction of relationship

.big-text-font[
- Lastly, we can describe the direction of relationship between two variables
- If one variable increases as the other variable increases --> *positive relationship*
- If one variable decreases as the other variable increases --> *negative relationship*
]

---

# Positive relationship

.left-code[
```{r positive-relationship, eval=FALSE}
x <- c(1:50)
y <- c(1:50) + 
  rnorm(50, 
        mean = 0, 
        sd = 0.2)
plot(x, y)
abline(a = 0, 
       b = 1, 
       col = "red")
```
]

.right-plot[
```{r positive-relationship-out, ref.label="positive-relationship", echo=FALSE}
```
]

---

# Negative relationship

.left-code[
```{r negative-relationship, eval=FALSE}
x <- c(1:50)
y <- c(50:1) + 
  rnorm(n = 50, 
        mean = 0, 
        sd = 1)
plot(x, y)
abline(a = 50, 
       b = -1, 
       col = "red")
```
]

.right-plot[
```{r negative-relationship-out, ref.label="negative-relationship", echo=FALSE}
```
]

---

# Putting it all together

.big-text-font[
- We can describe the relationship between two variables by combining descriptions for the linearity, strength, and direction of a relationship
- Example: the relationship between height and weight is a moderate, positive and linear relationship. 
- Example: the relationship between temperature and air conditioning use is a strong, negative, and linear relationship. 
]

